pipeComp, a general framework for the evaluation of computational pipelines, reveals performant single cell RNA-seq preprocessing tools

### 2020-07-03 bmc_article.tex
Updates from Overleaf

The impact of the dimensionality reduction methods tested was far greater than that of normalization or feature selection (Supplementary Figure 23). Overall, weighting the components by their explained variance tended to increase silhouette width (Supplementary Figure 23A) and, in most datasets, clustering accuracy (Supplementary Figure 23B). GLM-PCA tended to increase the average silhouette width of already well-defined subpopulations, but \textit{Seurat}'s PCA procedure however proved superior on all metrics. {\color{red} Like GLM-PCA, \textit{scVI}'s Linearly Decoded (LD) data \cite{SvenssonLDVAE2020} and latent space do not explicitly rely on normalized counts. Their performance was however worse than \textit{Seurat}'s PCA in view of the silhouette width (Supplementary Figure 15). In terms of clustering accuracy \textit{scVI}'s latent space performed better that its LD but was nevertheless less accurate than PCA-based alternatives on the datasets with a very low amount of cells (Koh and Kumar) and harboring many subpopulations (simMix1 and Zhengmix8eq).}

@@

The greater importance of normalization, dimensionality and feature selection over filtering steps is also visible in the clustering of (Figure \ref{fig:overall}). The top methods (bottom of Figure \ref{fig:overall}) confirm the results of the previous sections, highlighting especially \textit{sctransform}, deviance-based feature selection, and \textit{maxLikGlobal} dimensionality estimates as top methods. Interestingly, while the combination of \textit{sctransform} with \textit{Seurat} clustering was particularly performant for most datasets, for the simulated `simMix1` dataset instead standard \textit{Seurat} normalization combined with \textit{scran}-based clustering proved superior.}

@@

{\color{red} For the molecular cross-validation approach, we relied on a R implementation by Matthew Young (\url{https://github.com/constantAmateur/MCVR}), with minor modifications (see wrapper in the \textit{pipeComp} package). Briefly, since there is no a priori relationship between the dimensionality of the count data and that of the PCA based on the normalized data, MCVR does not use the original approach based Poisson loss, but the mean squared error of the normalized data, and uses balanced 50-50 splits. Because of the 50\% split, dimensionality may be underestimated, and titration did suggest that was the case for some of our datasets.}

@@

The evaluation of \textit{scVI} algorithm was performed separate to the other pipeline combinations as the model training on our reference datasets was especially resource-consuming {\color{red} and follows a different flow from the other pipelines. In particular, for datasets with fewer cells than genes, \textit{scVI} requires a feature selection to be performed prior to the normalization and imputation steps; furthermore, its dimensional reduction steps rely directly on raw counts. We therefore evaluated it separately and compared it to a representative subset of pipelines using \textit{sctransform} normalization and \textit{Seurat}'s PCA dimension reduction method. The resulting evaluation is available in the Supplementary Figure 13.

Using \textit{scVI}, we considered as imputed values the mean of the negative binomial distribution modeling the counts (accessed through the \textit{imputation} method). Following \textit{scVI}'s online documentation, we used the \textit{VAE} model, using a train size of 1, a training on 400 epochs and a learning rate of 0.001. As a major limitation of \textit{scVI} is the training on dataset with low gene to cell ratio, the features were subsampled to 1000 using the \textit{subsample\_genes} function, which resembles the \textit{Seurat} V3 VST gene selection method. All other parameters were left unmodified. The normalized values were computed from the normalized mean of the negative binomial distribution modeling the counts (accessed through the \textit{get\_sample\_scale} method).

@@

Filtering was performed with edgeR's \texttt{filterByExpr} function, changing the \texttt{min.count} argument. For \textit{vstsva}, we first applied the variance-stabilizing transformation from \textit{DESeq2}\citep{loveModeratedEstimation2014} before applying \texttt{sva::sva}. \textit{edgeR.QLF} refers to the quasi-likelihood version \citep{lundDetectingDifferential2012} of \textit{edgeR}'s fit. The exact wrappers used can be found in the package at \url{https://github.com/plger/pipeComp/blob/master/inst/extdata/dea\_wrappers.R}.
}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Backmatter begins here                   %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

The impact of the dimensionality reduction methods tested was far greater than that of normalization or feature selection (Supplementary Figure 23). Overall, weighting the components by their explained variance tended to increase silhouette width (Supplementary Figure 23A) and, in most datasets, clustering accuracy (Supplementary Figure 23B). GLM-PCA tended to increase the average silhouette width of already well-defined subpopulations, but \textit{Seurat}'s PCA procedure however proved superior on all metrics. {\color{red} Like GLM-PCA, \textit{scVI}'s Linearly Decoded (LD) data \cite{SvenssonLDVAE2020} and latent space do not explicitly rely on normalized counts. Their performance was however worse than \textit{Seurat}'s PCA in view of the silhouette width (Supplementary Figure 15). In terms of clustering accuracy, \textit{scVI}'s latent space performed better that its LD but was nevertheless less accurate than PCA-based alternatives on the datasets with a very low amount of cells (Koh and Kumar) and harboring many subpopulations (simMix1 and Zhengmix8eq).}

@@

The greater importance of normalization, dimensionality and feature selection over filtering steps is also visible in the clustering (Figure \ref{fig:overall}). The top methods (bottom of Figure \ref{fig:overall}) confirm the results of the previous sections, highlighting especially \textit{sctransform}, deviance-based feature selection, and \textit{maxLikGlobal} dimensionality estimates as top methods. Interestingly, while the combination of \textit{sctransform} with \textit{Seurat} clustering was particularly performant for most datasets, for the simulated `simMix1` dataset,  standard \textit{Seurat} normalization combined with \textit{scran}-based clustering proved superior.}

@@

{\color{red} For the molecular cross-validation approach, we relied on a R implementation by Matthew Young (\url{https://github.com/constantAmateur/MCVR}), with minor modifications (see wrapper in the \textit{pipeComp} package). Briefly, since there is no \textit{a priori} relationship between the dimensionality of the count data and that of the PCA based on the normalized data, MCVR does not use the original approach based Poisson loss, but the mean squared error of the normalized data, and uses balanced 50-50 splits. Because of the 50\% split, dimensionality may be underestimated, and titration did suggest that was the case for some of our datasets.}

@@

The evaluation of \textit{scVI} algorithm was performed separate to the other pipeline combinations since the model training on our reference datasets was especially resource-consuming {\color{red} and follows a different flow from the other pipelines. In particular, for datasets with fewer cells than genes, \textit{scVI} requires a feature selection to be performed prior to the normalization and imputation steps; furthermore, its dimensional reduction steps rely directly on raw counts. We therefore evaluated it separately and compared it to a representative subset of pipelines using \textit{sctransform} normalization and \textit{Seurat}'s PCA dimension reduction method. The resulting evaluation is available in the Supplementary Figure 13.

Using \textit{scVI}, we considered as imputed values the mean of the negative binomial distribution modeling the counts (accessed through the \textit{imputation} method). Following \textit{scVI}'s online documentation, we used the \textit{VAE} model, using a train size of 1, a training on 400 epochs and a learning rate of 0.001. As a major limitation of \textit{scVI} is the training on dataset with low gene to cell ratio, the features were subsampled to 2000 using the \textit{subsample\_genes} function, which resembles the \textit{Seurat} V3 VST gene selection method. All other parameters were left unmodified. The normalized values were computed from the normalized mean of the negative binomial distribution modeling the counts (accessed through the \textit{get\_sample\_scale} method).

@@

Filtering was performed with edgeR's \texttt{filterByExpr} function, changing the \texttt{min.count} argument. For \textit{vstsva}, we first applied the variance-stabilizing transformation from \textit{DESeq2} \citep{loveModeratedEstimation2014} before applying \texttt{sva::sva}. \textit{edgeR.QLF} refers to the quasi-likelihood version \citep{lundDetectingDifferential2012} of \textit{edgeR}'s fit. The exact wrappers used can be found in the package at \url{https://github.com/plger/pipeComp/blob/master/inst/extdata/dea\_wrappers.R}.
}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Backmatter begins here                   %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
